---
title: "Session 4: Homework 2"
author: "Group 6: Sonakshi Gupta, Jean Francois Peters, Wybe Harms, Drishti Goyal, Suzy Wang, Zezhou Tang"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(wbstats)
library(countrycode)
library(patchwork)
library(gganimate)
```

# Climate change and temperature anomalies

If we wanted to study climate change, we can find data on the *Combined
Land-Surface Air and Sea-Surface Water Temperature Anomalies* in the
Northern Hemisphere at [NASA's Goddard Institute for Space
Studies](https://data.giss.nasa.gov/gistemp). The [tabular data of
temperature anomalies can be found
here](https://data.giss.nasa.gov/gistemp/tabledata_v4/NH.Ts+dSST.txt)

To define temperature anomalies you need to have a reference, or base,
period which NASA clearly states that it is the period between
1951-1980.

```{r weather_data, cache=TRUE}

weather <- 
  read_csv("https://data.giss.nasa.gov/gistemp/tabledata_v4/NH.Ts+dSST.csv", 
           skip = 1, 
           na = "***")

```

For each month and year, the dataframe shows the deviation of
temperature from the normal (expected).

You have two objectives in this section:

1.  Select the year and the twelve month variables from the `weather`
    dataset. We do not need the others (J-D, D-N, DJF, etc.) for this
    assignment. Hint: use `select()` function.

2.  Convert the dataframe from wide to 'long' format. Hint: use
    `gather()` or `pivot_longer()` function. Name the new dataframe as
    `tidyweather`, name the variable containing the name of the month as
    `month`, and the temperature deviation values as `delta`.

```{r tidyweather}

tidyweather <- weather %>%
  select(1:13) %>%
  pivot_longer(
    cols = 2:13,
    names_to = "Month", 
    values_to = "delta")

tidyweather

```

## Plotting Information

Let us plot the data using a time-series scatter plot, and add a
trendline. To do that, we first need to create a new variable called
`date` in order to ensure that the `delta` values are plot
chronologically.

```{r scatter_plot }

tidyweather <- tidyweather %>%
  mutate(date = ymd(paste(as.character(Year), Month, "1")),
         month = month(date, label=TRUE),
         year = year(date))

ggplot(tidyweather, aes(x=date, y = delta))+
  geom_point()+
  geom_smooth(color="red") + #plot the trend line
  theme_bw() +
  labs (
    title = "Weather Anomalies",
    x = "Date",
    y = "Delta"
  )

```

Is the effect of increasing temperature more pronounced in some months?

-   The above graph shows a clear seasonal affect over the years.
    However, if we look at the graph below, it shows the deviation per
    month have very similar trend lines, reflective of the above graph.

```{r facet_wrap}

ggplot(tidyweather, aes(x=date, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  facet_wrap(~month) + #analyse over a each month
  theme_bw() +
  labs (
    title = "Weather Anomalies per Month",
    x = "Year",
    y = "Delta"
  )

```

It is sometimes useful to group data into different time periods to
study historical data. For example, we often refer to decades such as
1970s, 1980s, 1990s etc. to refer to a period of time. NASA calcuialtes
a temperature anomaly, as difference form the base periof of 1951-1980.
The code below creates a new data frame called `comparison` that groups
data in five time periods: 1881-1920, 1921-1950, 1951-1980, 1981-2010
and 2011-present.

```{r intervals}

comparison <- tidyweather %>% 
  filter(Year>= 1881) %>%     #remove years prior to 1881
  #create new variable 'interval', and assign values based on criteria below:
  mutate(interval = case_when(
    Year %in% c(1881:1920) ~ "1881-1920",
    Year %in% c(1921:1950) ~ "1921-1950",
    Year %in% c(1951:1980) ~ "1951-1980",
    Year %in% c(1981:2010) ~ "1981-2010",
    TRUE ~ "2011-present"
  ))

```

Now that we have the `interval` variable, we can create a density plot
to study the distribution of monthly deviations (`delta`), grouped by
the different time periods we are interested in.

```{r density_plot}

comparison %>%
  group_by(interval) %>%
  ggplot(aes(
    delta, 
    fill = interval, #fill color by the different intervals
    color = interval)) +
  geom_density(alpha=1/3) +
  theme_bw() +
  labs(
    title = "Density Plot for Distribution of Monthly Temperature Deviations",
    x = "Delta",
    y = NULL
  )

```

So far, we have been working with monthly anomalies. However, we might
be interested in average annual anomalies.

```{r averaging }

#creating yearly averages
average_annual_anomaly <- tidyweather %>% 
  group_by(Year) %>%   #grouping data by Year
  
  # creating summaries for mean delta 
  # use `na.rm=TRUE` to eliminate NA (not available) values 
  summarise(mean_delta = mean(delta, na.rm = TRUE))

#plotting the data:
ggplot(average_annual_anomaly, 
       aes(x = Year,
           y = mean_delta)) +
  geom_point() +
  #Fit the best fit line, using LOESS method
  geom_smooth(method = "loess", se = TRUE) +
  #change theme to theme_bw() to have white background + black frame around plot
  theme_bw() +
  labs(
    title = "Average Annual Anomslies",
    x = "Year",
    y = "Average Delta"
  )
                        

```

## Confidence Interval for `delta`

[NASA points out on their
website](https://earthobservatory.nasa.gov/world-of-change/decadaltemp.php)
that

> A one-degree global change is significant because it takes a vast
> amount of heat to warm all the oceans, atmosphere, and land by that
> much. In the past, a one- to two-degree drop was all it took to plunge
> the Earth into the Little Ice Age.

Your task is to construct a confidence interval for the average annual
delta since 2011, both using a formula and using a bootstrap simulation
with the `infer` package. Recall that the dataframe `comparison` has
already grouped temperature anomalies according to time intervals; we
are only interested in what is happening between 2011-present.

```{r, calculate_CI_using_formula}

formula_ci <- comparison %>% 
  
  # choose the interval 2011-present
  filter(interval=="2011-present") %>%
  # calculate summary statistics for temperature deviation (delta) 
  summarise(
    # calculate mean, SD, count, SE, lower/upper 95% CI 
    mean_delta = mean(delta, na.rm = TRUE),
    sd_delta = sd(delta, na.rm = TRUE),
    count = n(),
    t_critical = qt(0.975, count-1),
    se_delta = sd_delta/sqrt(count),
    moe = t_critical*se_delta,
    delta_lower = mean_delta - moe,
    delta_upper = mean_delta + moe
  )

#print out formula_CI
formula_ci

```

```{r, calculate_CI_using_bootstrap}

library(infer)

set.seed(1234)

ci_bootstrap <- comparison %>%
  filter(interval=="2011-present") %>%  # choose the interval 2011-present
  infer::specify(response = delta) %>%
  infer::generate(reps = 1000, type = "bootstrap") %>%
  infer::calculate(stat = "mean")

ci_percentile <- ci_bootstrap %>%
  infer::get_confidence_interval(level = .95, type = "percentile")

ci_percentile
  
```

> What is the data showing us? Please type your answer after (and
> outside!) this blockquote. You have to explain what you have done, and
> the interpretation of the result. One paragraph max, please!

We used two methods: the formula and the bootstrap method, to calculate
the confidence interval of the delta which is N(1.02,1.11). This tells
us that with the given mean and sd, we are 95% certain that the true
mean delta falls between 1.02 and 1.11.

# Biden's Approval Margins

As we saw in class, fivethirtyeight.com has detailed data on [all polls
that track the president's
approval](https://projects.fivethirtyeight.com/biden-approval-ratings)

```{r, cache=TRUE}
approval_polllist <- read_csv('https://projects.fivethirtyeight.com/biden-approval-data/approval_polllist.csv') 

# Use `lubridate` to fix dates, as they are given as characters.
approval_polllist <- approval_polllist %>%
  mutate(
    modeldate = mdy(modeldate),
    startdate = mdy(startdate),
    enddate = mdy(enddate),
    createddate = mdy(createddate)
  )

glimpse(approval_polllist)

```

## Create a plot

What I would like you to do is to calculate the average net approval
rate (approve- disapprove) for each week since he got into office. I
want you plot the net approval for each week in 2022, along with its 95%
confidence interval. There are various dates given for each poll, please
use `enddate`, i.e., the date the poll ended.

```{r trump_margins,  out.width="100%"}

approval_polllist %>%
  filter(year(enddate) == 2022) %>% #filtering for year 2022
  mutate(Week = week(enddate)) %>% #calculating week number
  group_by(subgroup, Week) %>%
  mutate(rating = approve - disapprove) %>% #calculating net approval rate
  
  #create confidence interval at 95% confidence
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    sd_rating = sd(rating, na.rm = TRUE),
    count = n(),
    t_critical = qt(.975,count-1),
    se_rating = sd_rating/sqrt(count),
    moe = t_critical * se_rating,
    rating_low = mean_rating - moe,
    rating_high = mean_rating + moe
  ) %>%
  
  # plot the graph
  ggplot(aes(x=Week, color = subgroup, fill = subgroup)) +
  geom_line(aes(y=rating_low), size = 1.2) +
  geom_line(aes(y=rating_high), size = 1.2) +
  geom_line(aes(y=mean_rating)) +
  facet_wrap(~subgroup, ncol = 1) +
  #to fill the area between the confidence interval lines
  geom_ribbon(aes(
    ymin = rating_low,
    ymax = rating_high
  ),fill = "orange", alpha = 1/3) +
  theme_bw() +
  theme(legend.position = "none") +
  
  #format the plot
  labs(
    title = "Biden's Net Approval Ratings in 2022",
    subtitle = "Weekly Data, Approve - Disapprove",
    x = "Week in 2022",
    y = NULL
  )

```

# Challenge 1: Excess rentals in TfL bike sharing

Recall the TfL data on how many bikes were hired every single day. We
can get the latest data by running the following

```{r, get_tfl_data, cache=TRUE}
url <- "https://data.london.gov.uk/download/number-bicycle-hires/ac29363e-e0cb-47cc-a97a-e216d900a6b0/tfl-daily-cycle-hires.xlsx"

# Download TFL data to temporary file
httr::GET(url, write_disk(bike.temp <- tempfile(fileext = ".xlsx")))

# Use read_excel to read it as dataframe
bike0 <- read_excel(bike.temp,
                   sheet = "Data",
                   range = cell_cols("A:B"))

# change dates to get year, month, and week
bike <- bike0 %>% 
  clean_names() %>% 
  rename (bikes_hired = number_of_bicycle_hires) %>% 
  mutate (year = year(day),
          month = lubridate::month(day, label = TRUE),
          week = isoweek(day))
```

We can easily create a facet grid that plots bikes hired by month and
year since 2015

```{r tfl_month_year_grid, out.width="100%"}

#create facet_grid
bike %>%
  filter(year > 2014) %>%
  ggplot(aes(bikes_hired)) +
  geom_density() +
  facet_grid(vars(year),vars(month)) +
  theme_bw() +
  labs(
    title = "Distribution of Bikes Hired per month since 2015",
    x = "Distribution of Bikes Hired",
    y = NULL
  ) +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())


```

However, the challenge I want you to work on is to reproduce the
following two graphs.

```{r tfl_absolute_monthly_change, out.width="100%"}

# calculate expected bikes hired using 2016-2019
expected <- bike %>%
  filter(year >= 2016 & year <=2019) %>%
  group_by(month) %>%
  mutate(expected_monthly = mean(bikes_hired)) %>%
  select(expected_monthly, month)

# calculate actual monthly average of bikes
actual <- bike %>%
  filter(year >= 2017) %>%
  group_by(year, month) %>%
  mutate(actual_monthly = mean(bikes_hired))

# combined the two dfs  
final <- left_join(actual, expected, by = 'month') %>%  
  mutate(diff = actual_monthly - expected_monthly)

final %>% 
  ggplot(aes(x=month, month = 1)) +
    geom_line(aes(y = expected_monthly, group =1), color="#0096FF", size = 1.5) +
    geom_line(aes(y = actual_monthly, group = 1)) +
  # geom_ribbon to add green
    geom_ribbon(aes(
      ymax = expected_monthly, 
      ymin = pmin(diff, 0) + expected_monthly), 
      fill = "#C70039", 
      alpha = 0.4,
      group = 1) +
  # geom_ribbon to add red
    geom_ribbon(aes(
      ymax = actual_monthly, 
      ymin = actual_monthly - pmax(diff,0)), 
      fill = "#007500", 
      alpha = 0.4,
      group = 1) +
  # facet by year
    facet_wrap(~year)+
    labs(title = "Month changes in TfL bike rentals", 
         subtitle = "Change from monthly average shown in blue and calculated between 2016-2019", 
         caption = "Source: TfL, London Data Store", 
         x = "Months",
         y = "Bike rentals")+
    theme_bw()
```

The second one looks at percentage changes from the expected level of
weekly rentals. The two grey shaded rectangles correspond to Q2 (weeks
14-26) and Q4 (weeks 40-52).

```{r tfl_percent_change, out.width="100%"}

#calculate actual bikes hired
bike_week1 <- bike %>%
  filter(year > 2016) %>%
  group_by(year, week) %>%
  mutate(mean_bikes = mean(bikes_hired))

#calculate expected hired bikes
bike_week2 <- bike %>%
  filter(year >=2016, year <= 2019) %>%
  group_by(week) %>%
  mutate(expected_hired = mean(bikes_hired))

# join the two tables and calculate percent_change
bikes_week <- bike_week1 %>%
  full_join(bike_week2, by="week") %>%
  mutate(
    pct_change = (mean_bikes - expected_hired)/expected_hired,
    color_id = pct_change > 0)

#create plot
ggplot(bikes_week,
       aes(x = week)) +
  annotate("rect",fill ="grey",alpha = 0.5,xmin = 14,xmax = 26, ymin =-Inf,ymax =Inf) +
  annotate("rect",fill ="grey",alpha = 0.5,xmin = 40,xmax = 52, ymin =-Inf,ymax =Inf) +
  geom_line(aes(y = pct_change)) +
  facet_wrap(~year.x) +
  
  #fill the pct_change under 0%
  geom_ribbon(
    aes(ymin = 0,
        ymax = pmax(0,pct_change),
        fill = "red"),
    alpha = .4) +
  
  #fill the pct_change over 0%
  geom_ribbon(
    aes(ymin = pmin(0, pct_change),
        ymax = 0,
        fill = "green"),
    alpha = .4) +
  
  #create geom_rug()
  geom_rug(aes(color=color_id),sides = "b") +
  
  
  #format the plot
  labs(
    title = "Weekly changes in TfL bike rentals",
    subtitle = "% changes from weekly averages calculated between 2016-2019",
    x = "Week",
    y = NULL
  ) +
  theme_bw() +
  theme(legend.position = "none")


```

Should you use the mean or the median to calculate your expected
rentals? Why?

-   We decided to use mean, because the data showed seasonal trends and
    there were not many clear outliers, which would affect the data.

# Challenge 2: Share of renewable energy production in the world

The National Bureau of Economic Research (NBER) has a a very interesting
dataset on the adoption of about 200 technologies in more than 150
countries since 1800. This is the[Cross-country Historical Adoption of
Technology (CHAT)
dataset](https://www.nber.org/research/data/cross-country-historical-adoption-technology).

```{r,load_technology_data}

technology <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-19/technology.csv')

#get all technologies
labels <- technology %>% 
  distinct(variable, label)

# Get country names using 'countrycode' package
technology <- technology %>% 
  filter(iso3c != "XCD") %>% 
  mutate(iso3c = recode(iso3c, "ROM" = "ROU"),
         country = countrycode(iso3c, origin = "iso3c", destination = "country.name"),
         country = case_when(
           iso3c == "ANT" ~ "Netherlands Antilles",
           iso3c == "CSK" ~ "Czechoslovakia",
           iso3c == "XKX" ~ "Kosovo",
           TRUE           ~ country))

#make smaller dataframe on energy
energy <- technology %>% 
  filter(category == "Energy")

# download CO2 per capita from World Bank using {wbstats} package
# https://data.worldbank.org/indicator/EN.ATM.CO2E.PC
co2_percap <- wb_data(country = "countries_only", 
                      indicator = "EN.ATM.CO2E.PC", 
                      start_date = 1970, 
                      end_date = 2022,
                      return_wide=FALSE) %>% 
  filter(!is.na(value)) %>% 
  #drop unwanted variables
  select(-c(unit, obs_status, footnote, last_updated))

# get a list of countries and their characteristics
# we just want to get the region a country is in and its income level
countries <-  wb_cachelist$countries %>% 
  select(iso3c,region,income_level)

```

First, produce a graph with the countries with the highest and lowest %
contribution of renewables in energy production. This is made up of
`elec_hydro`, `elec_solar`, `elec_wind`, and `elec_renew_other`. You may
want to use the *patchwork* package to assemble the two charts next to
each other.

```{r min-max_renewables, out.width="100%"}

library(patchwork)

x <- energy %>%
  pivot_wider(names_from=variable, values_from=value) %>% #used pivot_wider to get a value for each variable
  filter(year==2019) %>%
  group_by(country) %>%
  summarise(elec_hydro = sum(elec_hydro, na.rm=TRUE),
           elec_solar = sum(elec_solar, na.rm=TRUE),
           elec_wind = sum(elec_wind, na.rm=TRUE),
           elec_renew = sum(elec_renew_other, na.rm=TRUE),
           elecprod = sum(elecprod, na.rm=TRUE)) %>%
  mutate(percentage = (elec_hydro+elec_solar+elec_wind+elec_renew)/elecprod, #calc the total pct
         # dealing with na and infinites in the calculation
         percentage = if_else(is.na(percentage),0,percentage), 
         percentage = if_else(is.infinite(percentage),0,percentage)) %>% 
  filter(percentage>0) #filter for all positive and non-zero percentages

top_20 <- x%>%
  mutate(country=fct_reorder(country,percentage)) %>% #reordering country by percentage
  arrange(desc(percentage)) %>%
  slice_max(country,n=20) %>% #deriving the top 20
  #plot the graph
  ggplot(aes(percentage,country)) + 
  geom_col() +
  theme_bw() +
  scale_x_continuous(labels = scales::percent_format(scale = 100))

bottom_20 <- x%>%
  mutate(country=fct_reorder(country,percentage)) %>% #reordering country by percentage
  arrange(percentage) %>%
  slice_min(country,n=20) %>% #deriving the bottom 20
  ggplot(aes(percentage,country)) + 
  geom_col() +
  theme_bw() +
  scale_x_continuous(labels = scales::percent_format(scale = 100))

plot20 <- (top_20 + bottom_20) + #combining the two graphs
  plot_annotation("Highest and Lowest % of Renewables in Energy Production")
plot20

```

Second, you can produce an animation to explore the relationship between
CO2 per capita emissions and the deployment of renewables. As the % of
energy generated by renewables goes up, do CO2 per capita emissions seem
to go down?

-   As we can see from the graph the lower income groups have almost
    negligible CO2 per capita and thus, we cannot observe any
    significant pattern. However, for the higher income groups, we can
    see that over the years the renewable energy consumption has been
    increasing with reducing CO2 per capita. This can also be because,
    usage of renewable energy requires larger upfront capital which the
    higher income groups have.

```{r CO2perCapita}

y <- co2_percap %>%
  inner_join(countries, iso3c=iso3c) %>%
  rename(value_co2=value) %>%
  mutate(date=as.integer(date)) %>%
  left_join(energy, by=c('country'='country','date'='year'))

z <- y %>%
  pivot_wider(names_from=variable, values_from=value) %>%
  group_by(date,country) %>%
  summarise(elec_hydro=sum(elec_hydro, na.rm=TRUE),
           elec_solar=sum(elec_solar, na.rm=TRUE),
           elec_wind= sum(elec_wind, na.rm=TRUE),
           elec_renew=sum(elec_renew_other, na.rm=TRUE),
           elecprod=sum(elecprod, na.rm=TRUE),
           value_co2=sum(value_co2),
           income_level=income_level) %>%
  mutate(percentage=(elec_hydro+elec_solar+elec_wind+elec_renew)/elecprod,
         percentage = if_else(is.infinite(percentage),0,percentage)) %>%
  pivot_longer(starts_with("elec"), names_to="variable", values_to = "value")

ggplot(z, aes(percentage, value_co2, color=income_level)) +
  geom_point() +
  facet_wrap(~income_level) +
  labs(title = 'Year: {frame_time}', 
       x = '% renewables', 
       y = 'CO2 per cap') +
  transition_time(as.integer(date)) +
  scale_x_continuous(labels = scales::percent_format(scale = 100)) +
  ease_aes('linear') + 
  theme(legend.position = "none")




```

# Deliverables

As usual, there is a lot of explanatory text, comments, etc. You do not
need these, so delete them and produce a stand-alone document that you
could share with someone. Knit the edited and completed R Markdown file
as an HTML document (use the "Knit" button at the top of the script
editor window) and upload it to Canvas.

# Details

-   Who did you collaborate with: **Group 6 - Sonakshi Gupta, Jean
    Francois Peters, Wybe Harms, Drishti Goyal, Suzy Wang, Zezhou Tang**
-   Approximately how much time did you spend on this problem set: **6
    hrs 20 min 33 seconds**
-   What, if anything, gave you the most trouble: **Creating coloured
    ribbons for the graphs**

**Please seek out help when you need it,** and remember the [15-minute
rule](https://mam202.netlify.app/syllabus/#the-15-minute-rule){target="_blank"}.
You know enough R (and have enough examples of code from class and your
readings) to be able to do this. If you get stuck, ask for help from
others, post a question on Slack-- and remember that I am here to help
too!

> As a true test to yourself, do you understand the code you submitted
> and are you able to explain it to someone else?

# Rubric

Check minus (1/5): Displays minimal effort. Doesn't complete all
components. Code is poorly written and not documented. Uses the same
type of plot for each graph, or doesn't use plots appropriate for the
variables being analyzed.

Check (3/5): Solid effort. Hits all the elements. No clear mistakes.
Easy to follow (both the code and the output).

Check plus (5/5): Finished all components of the assignment correctly
and addressed both challenges. Code is well-documented (both
self-documented and with additional comments as necessary). Used
tidyverse, instead of base R. Graphs and tables are properly labelled.
Analysis is clear and easy to follow, either because graphs are labeled
clearly or you've written additional text to describe how you interpret
the output.
